---
title: "Data Preparation"
author: Kamryn Mansfield
---

### Load Libraries
```{r}
library(tidyverse)
library(sf)
library(ggspatial)
library(gt)
library(leaflet)
```


## Pull in and clean data

```{r}
combined_hsis_folder <- "../1_data/HSIS Charlotte Data/combined_data/"

busstop_data <- read_csv(paste0(combined_hsis_folder, "/","combined_18_to_24_catsStops.csv"))
busstop_data <- busstop_data[,2:ncol(busstop_data)]

crash_data <- read_csv(paste0(combined_hsis_folder, "/","combined_18_to_24_crash.csv"))
crash_data <- crash_data[,2:ncol(crash_data)]

person_data <- read_csv(paste0(combined_hsis_folder, "/","combined_18_to_24_per.csv"))
person_data <- person_data[,2:ncol(person_data)]

unit_data <- read_csv(paste0(combined_hsis_folder, "/","combined_18_to_24_unit.csv"))
unit_data <- unit_data[,2:ncol(unit_data)]

```

## Combining Data
I noticed that some files have an extra digit at the end, so I selected all the files with common lat lon formats and made two plots. 

small: These datasets have 5 or 6 digits in the coordinates and seem to just be missing a zero at the end. 
```{r}
small_coords_stops <- busstop_data |>
  filter(orgnl_file %in% c("clt18catsStops.csv",
                           "clt19catsStops.csv",
                           "clt20catsStops.csv",
                           "clt22catsStops.csv")) |>
  mutate(X = X*10, Y = Y*10)

```

big: These datasets have 6 or 7 digits in the coordinates
```{r}
big_coords_stops <- busstop_data |>
  filter(orgnl_file %in% c("clt23catsStops.csv",
                           "clt24catsStops.csv",
                           "clt21catsStops.csv"))
```

this df has combined the data after adjusting the coordinates
```{r}
uniform_coords_df <- bind_rows(small_coords_stops, big_coords_stops)

# ggplot(uniform_coords_df) +
#   geom_point(aes(x = X, y = Y))

```

While the documentation just identifies x and y as latitude and longitude, an AI search helped me find that they are specifically from the North Carolina State Plane Coordinate System, EPSG:2264
```{r}
# 1. Turn coordinates into sf object (North Carolina State Plane Coordinate System, EPSG:2264)
busstops_sf <- st_as_sf(uniform_coords_df, coords = c("X", "Y"), crs = 2264)
# 2. Transform to Lat/Long (WGS84, EPSG:4326)
busstops_sf <- st_transform(busstops_sf, 4326)

```


Now I will read in the pedestrian bus-related crashes that I already organized
```{r}
pb_crashes_sf <- st_read("../1_data/NCDOT bus related crashes/ped_bus_crashes.geojson")
```

A quick view of the pb_crashed over the bus stop data. There are some that don't look like they are very close to bus stops. 
```{r}
ggplot(busstops_sf) +
  annotation_map_tile() +
  geom_sf(color = "black") +
  geom_sf(data = pb_crashes_sf, color = "red")
```

I'm just looking at the different variables connected to the bus stops that I can use in my analysis. 
```{r}
lapply(busstop_data, unique)
```

Adding variable to busstop data: ROUTES_PASSING
```{r}
busstops_sf <- busstops_sf |>
  mutate(ROUTES_PASSING = sapply(strsplit(ROUTES,","), length))
```

Getting the year for each stop
```{r}
busstops_sf <- busstops_sf |>
  mutate(YEAR = case_when(
    orgnl_file == "clt18catsStops.csv" ~ 2018,
    orgnl_file == "clt19catsStops.csv" ~ 2019,
    orgnl_file == "clt20catsStops.csv" ~ 2020,
    orgnl_file == "clt21catsStops.csv" ~ 2021,
    orgnl_file == "clt22catsStops.csv" ~ 2022,
    orgnl_file == "clt23catsStops.csv" ~ 2023,
    orgnl_file == "clt24catsStops.csv" ~ 2024
  )) |>
  select(!orgnl_file)
```

Finding the closest bus stop to the crash
```{r}

pb_crashes_sf$year <- pb_crashes_sf$CrashDate |> year()

pb_crashes_sf <- pb_crashes_sf |>
  mutate(busstop_year = case_when(
    year < 2018 ~ 2018,
    TRUE ~ year
  ))

# make a unique id for each bus stop and year
busstops_sf <- busstops_sf |>
  mutate(ID = as.numeric(paste0(STOPID, YEAR)))

# I found some duplicate rows that I can delete
busstops_sf |>
  filter(ID %in% c(607102023,607102024)) |> 
  View()

# filtering out duplicate rows
busstops_sf <- busstops_sf |>
  mutate(is_duplicate = duplicated(ID)) |>
  filter(is_duplicate == FALSE)

# making sure all IDs are unique
length(unique(busstops_sf$ID)) == nrow(busstops_sf)

matching_stops <- c()
crash_data_list <- list()
for (year_num in unique(busstops_sf$YEAR)){
  new_crash_sf <- pb_crashes_sf |>
  filter(busstop_year == year_num)
  
  new_stop_sf <- busstops_sf |>
  filter(YEAR == year_num)
  
  idx <- st_nearest_feature(new_crash_sf, new_stop_sf)
  
  matching_stops <- c(matching_stops,new_stop_sf$ID[idx])

  crash_data_list[[as.character(year_num)]] <- new_crash_sf
}

#delete after
# new_stop_sf |>
#   filter(stop_idx)
#   ggplot() +
#   geom_sf(color = "blue") +
#   geom_sf(data = new_crash_sf, color = "red")

pb_crashes_sf <- bind_rows(crash_data_list)
pb_crashes_sf$ID <- matching_stops

length(unique(pb_crashes_sf$ID)) == nrow(pb_crashes_sf)

```

Now that I know the closest stop id of each crash, I want to find the distance from the stop and I want to view the crashes and bus stops on a map. 
```{r}
busstops_crashes <- busstops_sf[match(pb_crashes_sf$ID, busstops_sf$ID),]

distances <- st_distance(pb_crashes_sf, busstops_crashes, by_element = TRUE)

pb_crashes_sf$dist_from_stop <- as.numeric(distances)

busstops_small <- busstops_crashes |>
  select(ID) |>
  mutate(category = "stop")

combined_to_plot <- pb_crashes_sf |>
  select(ID) |>
  mutate(category = "crash") |>
  bind_rows(busstops_small)


# 1. Define a palette for True/False values
# You can use hex codes or standard color names like "green" and "red"
pal <- colorFactor(palette = c("skyblue3", "red"), domain = c("stop", "crash"))

leaflet(combined_to_plot) %>%
  addTiles() %>%
  addCircleMarkers(
    # 2. Color points by the boolean column using the palette
    color = ~pal(category),
    # 3. Add a popup label (shows on click)
    popup = ~ID,
    # Optional: add a label (shows on hover)
    label = ~ID,
    fillOpacity = 0.7,
    radius = 4
  )

```

I want to add the crash data to the bus data frame, and then save it as a geojson file to use more easily in the main analysis.
```{r}
pb_no_geom <- st_drop_geometry(pb_crashes_sf)

combined_stops_crashes <- left_join(busstops_sf, pb_no_geom, by = "ID")
combined_stops_crashes$crash_year <- combined_stops_crashes$year
combined_stops_crashes$year <- NULL

st_write(combined_stops_crashes, "../1_data/NCDOT bus related crashes/combined_stops_crashes.geojson")

```
