---
title: Data Exploration
author: Kamryn Mansfield

---

```{r}
#| output: false

library(tidyverse)
library(sf)
library(ggspatial)
source("../R/functions.R")
```


Combine HSIS Data:

I made a function to combine the data, but they are too big for github. So I put folder they are saved to in the gitignore.

```{r}
#### COMBINE HSIS DATA ####
# run this once to combined the files

# The folder with the hsis data you want to combine
# hsis_folder <- "../1_data/HSIS Charlotte Data/"
#
# This functions combines the appropriate hsis files if you have multiple years
# combine_HSIS_files(hsis_folder)
```


```{r}
#### PULL IN AND CLEAN DATA ####
combined_hsis_folder <- "../1_data/HSIS Charlotte Data/combined_data/"

busstop_data <- read_csv(paste0(combined_hsis_folder, "/","combined_18_to_24_catsStops.csv"))
busstop_data <- busstop_data[,2:ncol(busstop_data)]
crash_data <- read_csv(paste0(combined_hsis_folder, "/","combined_18_to_24_crash.csv"))
crash_data <- crash_data[,2:ncol(crash_data)]
```

Making a data frame that will give true for each column that had no changes throughout the year
```{r}

stop_changes <- busstop_data |>
  group_by(StopID) |>
  summarise(x_unq = length(unique(x)) == 1,
            y_unq = length(unique(y)) == 1,
            Bench_unq = length(unique(Bench)) == 1,
            Shelter_unq = length(unique(Shelter)) == 1,
            routes_unq = length(unique(routes)) == 1,
            Direction_unq = length(unique(Direction)) == 1,
            Pole_Locat_unq = length(unique(Pole_Locat)) == 1,
            Sidewalk_unq = length(unique(Sidewalk)) == 1,
            Trans_P_unq = length(unique(Trans_P)) == 1,
            orgnl_file = length(unique(orgnl_file)) == 1)
```

it looks like there have been some changes throughout the year
```{r}

colSums(stop_changes)

```

looking at the unique stop ids
```{r}
stop_id_count <- busstop_data |>
  group_by(StopID) |>
  summarise(count = n(),
            file = paste(unique(orgnl_file), collapse = ", "))

stop_id_count |>
  filter(count == 2) |>
  select(file) |>
  unique()

busstop_data |>
  filter(orgnl_file == "clt24catsStops.csv") |>
  arrange(StopID)

```

When I plot lon lat points as is, they don't line up.
```{r}

busstop_data |>
  arrange(StopID) |>
  ggplot() +
  geom_point(aes(x = x, y = y, color = orgnl_file))


```

I noticed that some files have an extra digit at the end, so I selected all the files with common lat lon formats and made two plots. 
```{r}
busstop_data |>
  filter(orgnl_file %in% c("clt18catsStops.csv",
                           "clt19catsStops.csv",
                           "clt20catsStops.csv",
                           "clt22catsStops.csv")) |>
  ggplot() +
  geom_point(aes(x = x, y = y, color = orgnl_file))

busstop_data |>
  filter(orgnl_file %in% c("clt23catsStops.csv",
                           "clt24catsStops.csv",
                           "clt21catsStops.csv")) |>
  ggplot() +
  geom_point(aes(x = x, y = y, color = orgnl_file))
```

testing to see if I can get all stops on the same map

small: These datasets have 5 or 6 digits in the coordinates
```{r}
small_coords_stops <- busstop_data |>
  filter(orgnl_file %in% c("clt18catsStops.csv",
                           "clt19catsStops.csv",
                           "clt20catsStops.csv",
                           "clt22catsStops.csv")) |>
  mutate(x = x*10, y = y*10)

```

big: These datasets have 6 or 7 digits in the coordinates
```{r}
big_coords_stops <- busstop_data |>
  filter(orgnl_file %in% c("clt23catsStops.csv",
                           "clt24catsStops.csv",
                           "clt21catsStops.csv"))
```

this df has combined the data after adjusting the coordinates
```{r}
bound_stops <- bind_rows(small_coords_stops, big_coords_stops) |>
  mutate(lon = x / 10000,
         lat = y / 10000)

```

Now I just need to find out how to get the coordinates as an sf object to plot on the map correctly.
I am not done with this code chunk yet. I think I am doing something wrong with the crs.
```{r}
df_sf <- st_as_sf(bound_stops, coords = c("lat", "lon"), crs = 4326)

# 1. Turn coordinates into sf object (assuming CRS 32610 - UTM Zone 10N)
df_sf <- st_as_sf(bound_stops, coords = c("x", "y"), crs = 26917)
# 2. Transform to Lat/Long (WGS84, EPSG:4326)
df_longlat <- st_transform(df_sf, crs = 4326)

ggplot(df_sf) +
  annotation_map_tile() +
  geom_sf(aes(color = orgnl_file))


```

